The imprint of global magma oceans on exoplanet demographics

### 2023-04-01 src/tex/ms.tex
flesh out background, introduce Bioverse

% Main body
\section{Introduction}
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
\todo[inline]{``Points for the intro:
\bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and \nautilus\ are \\
Possibility of follow-up to measureme masses
''}

\begin{itemize}
    \item hot/warm planets are best studied
    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992)
    \item HZ concept.''(Kasting et al. 1993; Kopparapu et al. 2013, 2014)'' Relevance in the search for habitable planets and life. (Science gap:) Does it exist? Can it be tested? => statistically detect inner edge
    \item Bioverse
    \item Transit survey => PLATO
    \item Using PLATO as a baseline survey, we do A, B, C.
\end{itemize}

%\begin{note}
%Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
%Due to the inherent detection biases of the transit technique, most current well-probed small planets are much closer to their host star than Earth is to the Sun. CITE
%These bodies experience climatic conditions that are similar to the environment of the inner Solar System planets at early stages of their formation. CITE
%Studies of the geophysical state and evolution of warm exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
%
%A much regarded object of study are the divergent atmospheric evolutions of Earth and Venus.
%While both have accreted from the same mass reservoir and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces. "(Donahue et al. 1982; Kasting 1988; Hamano et al. 2013; Way et al. 2016; Kane et al. 2019; Way \& Del Genio 2020; Turbet et al. 2021)."
%
%
%
%
%
%
%
%
%
%\end{note}

>>>>

% Main body
\section{Introduction}
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
\todo[inline]{``Points for the intro:
\bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and \nautilus\ are \\
Possibility of follow-up to measureme masses
''}

\begin{itemize}
    \item hot/warm planets are best studied
    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992). Transition is sharp in instellation.
    \item HZ concept.''(Kasting et al. 1993; Kopparapu et al. 2013, 2014)'' Relevance in the search for habitable planets and life. (Science gap:) Does it exist? Can it be tested? => statistically detect inner edge
    \item Bioverse
    \item Transit survey => PLATO
    \item Using PLATO as a baseline survey, we do A, B, C.
\end{itemize}

\begin{note}
Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
Due to the inherent detection biases of the transit technique, most current well-probed small planets are much closer to their host star than Earth is to the Sun. CITE
These bodies experience climatic conditions that are similar to the environment of the inner Solar System planets at early stages of their formation. CITE
Studies of the geophysical state and evolution of warm exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.

A much regarded object of study are the divergent atmospheric evolutions of Earth and Venus.
While both have accreted from a similar mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021}. "(Donahue et al. 1982; Bézard, B. \& de Bergh, 2007; Way et al. 2016; Way \& Del Genio 2020; )."
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
%magma oceans...
However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits outgoing radiation and thus the release of excess heat~\citep{Ingersoll1969,Kasting1988,Wordsworth2022}.
This runaway greenhouse state can not only extend the magma ocean stage of a rocky planet to hundreds of \SI{}{\mega\year}, it can also lead to a loss of its entire water reservoir by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen. CITE
Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.
A water-dominated atmosphere of a planet in the runaway greenhouse state is significantly inflated~\citep{Turbet2019,Mousis2020}, increasing its transit radius by an amount mainly depending on the surface water content.
An Earth-like planet experiences a \SI{\sim 500}{\kilo\meter} radius increase upon crossing the threshold instellation, and its transit depth increases by \SI{10}{\ppm} to \SI{1000}{\ppm} depending on the spectral type of the host star~\citep{Turbet2019}.
... water incorporation...

The radiation-induced transition between a runaway greenhouse state and a temperate climate has been predicted to be a sharp threshold in instellation~\citep{Hamano2013}. CITE
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1853,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
Specifically, we assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes an abrupt change of planetary radii and bulk densities when ordered by receiving instellation.
Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
It consists of a sample generator that populates stars from the \gaia\ catalog CITE with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
Trying different instrumentation and survey designs, we try to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.

Our goal is to...
We then perform a parameter study to explore which trade-offs in the survey design of a \plato-like mission optimize the chance to detect and characterize the runaway greenhouse transition.

\end{note}
